Surrogate model of aligned-spin binary neutron star waveforms using Gaussian process regression

### 2017-11-22 paper/paper.tex
Changed Figs 1 and 2 and discussion. New surrogate with almost reviewed waveform model in progress.

In this section we describe a method for decomposing the waveform into smooth, slowly-varying functions and training a surrogate model for these functions. We will work with the frequency domain waveform $\tilde h(f; \bx)$. This is favorable for data analysis which is usually done in the frequency domain. It also allows us to extend the model down to arbitrarily low frequencies with analytic, Frequency-domain PN approximations. Unfortunately, Fourier transforming a finite length, discretely sampled waveform, leads to a surrogate with more noise. We will show below, however, that sufficient filtering can mostly solve this problem.

@@

Because the waveform $\tilde h(f; \bx)$ is an oscillatory function of $f$ and $\bx$, the waveform is usually decomposed into an amplitude $A(f; \bx)$ and phase $\Phi(f; \bx)$ as $\tilde h(f; \bx) = A(f; \bx) e^{i\Phi(f; \bx)}$. These functions are smoother, mostly monotonic functions of $f$ and $\bx$. This can be seen in Fig.~\ref{fig:h} which shows the initial training set of waveforms. \red{[[Actually, you might want to show the 51 waveforms with only $S_1$ varied. Then show them as a function of $f$ and another plot as a function of $S_1$.]]} Unfortunately, the amplitude and phase still span a wide range of values. The phase, for example, spans about $10^4$~rad. To avoid systematic errors in the tidal parameters, we need phase errors of $\lesssim 1$~rad over most of this frequency range, leading to a requirement on the fractional interpolation error of $\lesssim 10^{-4}$~rad. This is a difficult requirement to achieve in the 5-dimensional parameter space here.

\begin{figure}[htb]
\centering
\includegraphics[width=0.49\textwidth]{h_trainingset.png}
%\includegraphics[width=0.49\textwidth]{dh_trainingset.png}
\caption{Training set. Maybe instead show the waveform set which varies 1 parameter.}
\label{fig:h}
\end{figure}

For aligned-spin waveforms, we can solve this problem using that fact that the waveform is approximately known analytically from the stationary phase approximation to the post-Newtonian (PN) waveform known as TaylorF2 [[cite]] $\tilde h_{\rm F2}(f; \bx) = A_{\rm F2}(f; \bx) e^{i\Phi_{\rm F2} (f; \bx)}$. This allows us to write the EOB amplitude and phase in terms of the small residuals, $\Delta\ln(A)(f; \bx)$ and $\Delta\Phi(f; \bx)$, relative to TaylorF2:
\begin{align}
\Delta\ln(A)(f; \bx) &= \ln\left(\frac{ A(f; \bx) }{ A_{\rm F2}(f; \bx)}\right), \label{eq:ampres}\\
\Delta\Phi(f; \bx) & = \Phi(f; \bx) - \Phi_{\rm F2}(f; \bx), \label{eq:phaseres}
\end{align}
such that
\begin{align}
\tilde h(f; \bx) = \tilde h_{\rm F2}(f; \bx) e^{\Delta\ln(A)(f; \bx) + i  \Delta\Phi(f; \bx)}.
\label{eq:hdecomp}
\end{align}
%\begin{align}
%\tilde h(f; \bx) = A_{\rm F2}(f; \bx) e^{i\Phi_{\rm F2}(f; \bx)} e^{\Delta\ln(A)(f; \bx) + i  \Delta\Phi(f; \bx)}.
%\label{eq:hdecomp}
%\end{align}
These residuals are shown in Fig.~\ref{fig:dh}. For the amplitude residual, we use a log-ratio instead of a ratio because it guarantees that the amplitude of the reconstructed waveform in Eq.~\eqref{eq:hdecomp} will be positive even if interpolation error leads to a negative $\Delta\ln(A)(f; \bx)$. In addition, because the waveform amplitude spans several orders of magnitude after the merger frequency of each waveform, the log-ratio captures this behavior. Comparing Fig.~\ref{fig:dh} to Fig.~\ref{fig:h}, we find that for most of the frequency range, the range in $\Delta\Phi$ is a few orders of magnitude smaller than the range in $\Phi$.

\begin{figure}[htb]
\centering
%\includegraphics[width=0.49\textwidth]{h_trainingset.png}\\
\includegraphics[width=0.49\textwidth]{dh_trainingset.png}
\caption{Training set. Maybe instead show the waveform set which varies 1 parameter.}
\label{fig:dh}
\end{figure}

@@

@@

\subsection{Conditioning the training set waveforms}

@@

\red{[[Show plot (?) to validate that waveform is properly conditioned by plotting amplitude/phase as function of waveform parameters at specific frequencies and show that they're smooth functions?]]}

>>>>

In this section we describe how we decompose the waveform into smooth, slowly-varying functions and train a surrogate model for these functions. We will work with the frequency-domain waveform $\tilde h(Mf; \bx)$. This is favorable for data analysis which is usually done in the frequency domain. It also allows us to extend the model down to arbitrarily low frequencies with analytic, Frequency-domain PN approximations. Unfortunately, Fourier transforming a finite length, discretely sampled waveform leads to a surrogate with more noise. We will show below, however, that sufficient filtering can solve this problem.

@@

%Because the waveform $\tilde h(Mf; \bx)$ is an oscillatory function of $Mf$ and $\bx$, the waveform is usually decomposed into an amplitude $A(Mf; \bx)$ and phase $\Phi(Mf; \bx)$ as $\tilde h(Mf; \bx) = A(Mf; \bx) e^{i\Phi(Mf; \bx)}$. These functions are smoother, mostly monotonic functions of $Mf$ and $\bx$. This can be seen in Fig.~\ref{fig:h} which shows the initial training set of waveforms. \red{[[Actually, you might want to show the 51 waveforms with only $S_1$ varied. Then show them as a function of $f$ and another plot as a function of $S_1$.]]} Unfortunately, the amplitude and phase still span a wide range of values. The phase, for example, spans about $10^4$~rad. To avoid systematic errors in the tidal parameters, we need phase errors of $\lesssim 1$~rad over most of this frequency range, leading to a requirement on the fractional interpolation error of $\lesssim 10^{-4}$~rad. This is a difficult requirement to achieve in the 5-dimensional parameter space here.

%\begin{figure}[htb]
%\centering
%\includegraphics[width=0.49\textwidth]{h_trainingset.png}
%%\includegraphics[width=0.49\textwidth]{dh_trainingset.png}
%\caption{Training set. Maybe instead show the waveform set which varies 1 parameter.}
%\label{fig:h}
%\end{figure}

%\begin{figure}[htb]
%\centering
%%\includegraphics[width=0.49\textwidth]{h_trainingset.png}\\
%\includegraphics[width=0.49\textwidth]{dh_trainingset.png}
%\caption{Training set. Maybe instead show the waveform set which varies 1 parameter.}
%\label{fig:dh}
%\end{figure}

@@

Because the waveform $\tilde h(Mf; \bx)$ is an oscillatory function of $Mf$ and $\bx$, the waveform is usually decomposed into an amplitude $A(Mf; \bx)$ and phase $\Phi(Mf; \bx)$ as $\tilde h(Mf; \bx) = A(Mf; \bx) e^{i\Phi(Mf; \bx)}$. The amplitude and phase are smoother, mostly monotonic functions of frequency. This can be seen in Fig.~\ref{fig:hoff} where we show the waveforms for the 32 corners of the 5-dimensional parameter space. Unfortunately, the amplitude and phase still span a wide range of values. The phase, for example, spans about $10^4$~rad between waveforms with different parameters $\bx$ (see Fig.~\ref{fig:hoff}). To avoid systematic errors in the tidal parameters, we need phase errors of $\lesssim 1$~rad over most of this frequency range, leading to a requirement on the fractional interpolation error of $\lesssim 10^{-4}$~rad. This is a difficult requirement to achieve in the 5-dimensional parameter space here.

\begin{figure}[htb]
\centering
\includegraphics[width=0.49\textwidth]{hoff.png}
\caption{EOB waveforms $\tilde h(Mf; \bx)$ for the 32 corners of parameter space. The waveforms are filtered and Fourier transformed as described in Sec.~\ref{sec:condition}. Top two panels: the amplitude $A(Mf; \bx)$ and residual $\Delta\ln A(Mf; \bx)$ relative to TaylorF2 as defined in Eq.~\eqref{eq:ampres}. Bottom two panels: the phase $\Phi(Mf; \bx)$ and residual $\Delta\Phi(Mf; \bx)$ relative to TaylorF2 as defined in Eq.~\eqref{eq:phaseres}. Vertical dashed lines represent the frequency nodes $MF_j$ where the residuals are interpolated as functions of $\bx$ using Gaussian process regression.}
\label{fig:hoff}
\end{figure}

For aligned-spin waveforms, we can solve this problem using the fact that the waveform is approximately known analytically from the stationary phase approximation to the post-Newtonian (PN) waveform known as TaylorF2 [[cite]] $\tilde h_{\rm F2}(Mf; \bx) = A_{\rm F2}(Mf; \bx) e^{i\Phi_{\rm F2} (Mf; \bx)}$. This allows us to write the EOB amplitude and phase in terms of small residuals, $\Delta\ln(A)(Mf; \bx)$ and $\Delta\Phi(Mf; \bx)$, relative to TaylorF2:
\begin{align}
\Delta\ln(A)(Mf; \bx) &= \ln\left(\frac{ A(Mf; \bx) }{ A_{\rm F2}(Mf; \bx)}\right), \label{eq:ampres}\\
\Delta\Phi(Mf; \bx) & = \Phi(Mf; \bx) - \Phi_{\rm F2}(Mf; \bx), \label{eq:phaseres}
\end{align}
such that
\begin{align}
\tilde h(Mf; \bx) = \tilde h_{\rm F2}(Mf; \bx) e^{\Delta\ln(A)(Mf; \bx) + i  \Delta\Phi(Mf; \bx)}.
\label{eq:hdecomp}
\end{align}
These residuals are also shown in Fig.~\ref{fig:hoff}. For the amplitude residual, we use a log-ratio instead of a ratio because it guarantees that the amplitude of the reconstructed waveform (Eq.~\eqref{eq:hdecomp}) will be positive even if interpolation error leads to a negative value of $\Delta\ln(A)(Mf; \bx)$. In addition, because the waveform amplitude spans several orders of magnitude after the merger frequency of each waveform, the log-ratio captures this behavior. Comparing the phase $\Phi$ and phase residual $\Delta\Phi$ in Fig.~\ref{fig:hoff}, we find that the range in $\Delta\Phi$ is a few orders of magnitude smaller than the range in $\Phi$ except at very high frequencies. In Fig.~\ref{fig:dhofs}, we show the amplitude and phase residuals at three fixed frequencies as functions of the waveform parameter $S_1$. These quantities are smooth functions of $\bx$, except for the amplitude residual at very high frequencies where the amplitude is small and numerical noise dominates, implying that interpolation as functions of $\bx$ should work.

\begin{figure}[htb]
\centering
\includegraphics[width=0.49\textwidth]{dhofs.png}
\caption{Cross sections of the amplitude and phase residuals as functions of $S_1$ for the fixed frequencies $Mf=\{0.01, 0.03, 0.05\}$. The waveform parameters are $\bx = \{q, S_1, S_2, \Lambda_1, \Lambda_2\} = \{0.6, S_1, 0.2, 2000, 1000\}$. The waveforms are generated for 51 values of $S_1$ shown as dots. Also shown is the 1-dimensional cross section of the final surrogate using Gaussian process regression (solid black) as well as its estimated $1\sigma$ uncertainty (shaded gray region).}
\label{fig:dhofs}
\end{figure}

@@

\subsection{Conditioning the training set waveforms}
\label{sec:condition}

@@

To validate that the waveforms are sufficiently conditioned, we plot cross-sections of the residuals in Fig.~\ref{fig:dhofs} for fixed frequencies as functions of the waveform parameter $S_1$. The fact that the residuals are smooth functions of the waveform parameters indicates that most numerical noise has been removed. The exception is the noisy amplitude residual at high frequencies, $\Delta\ln(A)(Mf=0.05; \bx)$, where the amplitude is very small and dominated by numerical noise.
